---
output: github_document
editor_options: 
  chunk_output_type: console
bibliography: vignettes/PRDA.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(PRDAbeta)
```

# PRDAbeta: Prospective and Retrospective Design Analysis

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ClaudioZandonella/PRDAbeta?branch=develop&svg=true)](https://ci.appveyor.com/project/ClaudioZandonella/PRDAbeta)
[![Travis build status](https://travis-ci.org/ClaudioZandonella/PRDA_beta.svg?branch=develop)](https://travis-ci.org/ClaudioZandonella/PRDA_beta)
[![Codecov test coverage](https://codecov.io/gh/ClaudioZandonella/PRDAbeta/branch/develop/graph/badge.svg)](https://codecov.io/gh/ClaudioZandonella/PRDAbeta/branch/develop)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3630733.svg)](https://doi.org/10.5281/zenodo.3630733)
<hr>
<!-- badges: end -->

PRDAbeta allows performing a prospective or retrospective design analysis to evaluate inferential risks (i.e., power, Type M error, and Type S error) in a study considering Pearson's correlation between two variables or mean comparisons (one-sample, paired, two-samples, and Welch's t-test). 

For an introduction to design analysis and a general overview of the package see `vignettes("PRDAbeta_overview")`.
Examples for retrospective design analysis and prospective design analysis are provided in `vignettes("retrospective")` and `vignettes("prospective")` respectively.

## Installation

<!-- You can install the released version of PRDAbeta from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("PRDAbeta") -->
<!-- ``` -->

<!-- And the development version from [GitHub](https://github.com/) with: -->

You can install the development version from [GitHub](https://github.com/ClaudioZandonella/PRDAbeta/tree/develop) with:

``` r
# install.packages("devtools")
devtools::install_github("ClaudioZandonella/PRDAbeta",
                         ref = "develop",
                         build_vignettes = TRUE)
```


## The Package

PRDAbeta package can be used for Pearson's correlation between two variables or mean comparisons (one-sample, paired, two-samples, and Welch's t-test) considering a hypothetical value of *&rho;* or Cohen's *d* respectively. See `vignette("retrospective")` and `vignette("prospective")` to know how to set function arguments for the different effect types. 


### Functions

In PRDAbeta there are two main functions `retrospective()` and `prospective()`.

#### &#8226; `retrospective()`

Given the hypothetical population effect size and the study sample size, the function `retrospective()` performs a retrospective design analysis. According to the defined alternative hypothesis and the significance level, the inferential risks (i.e., Power level, Type M error, and Type S error) are computed together with the critical effect value (i.e., the minimum absolute effect size value that would result significant).

Consider a study that evaluated the correlation between two variables with a sample of 30 subjects. Suppose that according to the literature the hypothesized effect is *&rho;* = .25. To evaluate the inferential risks related to the study we use the function `retrospective()`.

```{r retrospective,}
retrospective(effect_size = .25, sample_n1 = 30,
              effect_type = "correlation", test_method = "pearson",
              seed = 2020)
```

In this case, the statistical power is almost 30% and the associated Type M error and Type S error are respectively around 1.80 and 0.003. That means, statistical significant results are on average an overestimation of 80% of the hypothesized population effect and there is a .3% of probability to obtain a statistically significant result in the opposite direction.

To know more about function arguments and further examples see the function documentation `?retrospective` and  `vignette("retrospective")`.

#### &#8226; `prospective()`

Given the hypothetical population effect size and the required power level, the function `prospective()` performs a prospective design analysis. According to the defined alternative hypothesis and the significance level, the required sample size is computed together with the associated Type M error, Type S error, and the critical correlation value (i.e., the minimum absolute effect size value that would result significant).  

Consider a study that will evaluate the correlation between two variables. Knowing from the literature that we expect an effect size of *&rho;* = .25, the function `prospective()` can be used to compute the required sample size to obtain a power of 80%. 
```{r prospective}
prospective(effect_size = .25, power = .80, 
            effect_type = "correlation", test_method = "pearson",
            display_message = FALSE, seed = 2020)
```

The required sample size is $n=126$, the associated Type M error is around 1.10 and the Type S error is approximately 0. 

To know more about function arguments and further examples see the function documentation `?prospective` and `vignette("prospective")`.

### Hypothetical effect size

The hypothetical population effect size can be defined as a single value according to previous results in the literature or experts indications. Alternatively, PRDAbeta allows users to specify a distribution of plausible values to account for their uncertainty about the hypothetical population effect size.  To know how to specify the hypothetical effect size according to a distribution and an example of application see `vignette("retrospective")`.

## Bugs and New Features

To propose a new feature or report a bug, please open an issue on [GitHub](https://github.com/ClaudioZandonella/PRDAbeta/issues).

## Citation

To cite PRDAbeta in publications use:

Claudio Zandonella Callegher, Massimiliano Pastore, Angela Andreella, Anna Vesely, Enrico Toffalini, Giulia Bertoldo, & Gianmarco Altoè. (2020).  PRDAbeta: Prospective and Retrospective Design Analysis (Version v0.1). Zenodo. http://doi.org/10.5281/zenodo.3630733

A BibTeX entry for LaTeX users is
```{}
@Misc{,
    title = {{PRDAbeta}: Prospective and Retrospective Design Analysis},
    author = {Claudio {Zandonella Callegher} and Massimiliano Pastore and Angela Andreella and Anna Vesely and Enrico Toffalini and Giulia Bertoldo and Gianmarco Altoè},
    year = {2020},
    publisher = {Zenodo},
    version = {v0.1},
    doi = {10.5281/zenodo.3630733},
    url = {https://doi.org/10.5281/zenodo.3630733},
  }
```

---
nocite: | 
  @altoeEnhancingStatisticalInference2020, @bertoldoDesigningStudiesEvaluating2020, @gelmanPowerCalculationsAssessing2014
...

## References

